---
title: "Alpha Diversity"
output:
  md_document:
     variant: markdown_github
  html_document:
    df_print: paged
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, include=TRUE)
```


## Resources
This draws from [phyloseq plot_bar tutorial](https://joey711.github.io/phyloseq/plot_bar-examples.html).

## Data
This tutorial uses the 10% Atacama subset data (note that for the demux and dada2 tutorial we used the 1% Atacama subset)

# Getting ready

First we load libraries.
```{r libraries, message=FALSE, warning=FALSE}
library(readr)
library(phyloseq)
```

```{r files_and_directories, include=FALSE}
# Directories
atacama.rds = "/data/tutorial_data/atacama_10pct.rds"

```

```{r}
atacama.ps = read_rds(atacama.rds)
print(atacama.ps)
```

# Visualize alpha-diversity
```{r warning=FALSE}
plot_richness(atacama.ps, x="TransectName", 
              measures=c("Observed", "Shannon", "Chao1"), color="TransectName") + theme_bw()
```

## Alpha-Diversity Boxplots
It is a bit hard to compare the two different transects because many points are overlapping, let's add a boxplot layer so we can compare the distribution of alpha-diversity values between the transects.
```{r warning=FALSE}
plot_richness(atacama.ps, x="TransectName", 
              measures=c("Observed", "Shannon", "Chao1"), color="TransectName") + 
              geom_boxplot() +
              geom_point() +
              theme_bw() 
```

## Alpha-Diversity as a function of other parameters
It also might be interesting to explore whether other parameters have an effect on alpha-diversity
```{r}
sample_variables(atacama.ps)
```

### Elevation
```{r warning=FALSE}
plot_richness(atacama.ps, x="Elevation", 
              measures=c("Observed", "Shannon", "Chao1"), color="TransectName") + theme_bw()
```

### Depth
```{r warning=FALSE}
plot_richness(atacama.ps, x="Depth", 
              measures=c("Observed", "Shannon", "Chao1"), color="TransectName") + theme_bw()
```

# Session Info
Always print `sessionInfo` for reproducibility!
```{r}
sessionInfo()
```

