---
output:
  html_document:
    df_print: paged
---
```{r global_options, include=FALSE}
# knitr::opts_chunk$set(eval=FALSE, echo=TRUE)
# knitr::opts_chunk$set(cache.path = cache.prefix)
```


# Getting ready

First we load libraries.
```{r libraries, message=FALSE, warning=FALSE}
library(readr)
library(phyloseq)
```

# Relative Abundance Plots
# Get Data
This tutorial uses the 10% Atacama subset data (note that for the demux and dada2 tutorial we used the 1% Atacama subset)

```{r files_and_directories, include=FALSE}
# Directories
atacama.rds = "/data/tutorial_data/atacama_1pct"

```

```{r}
atacama.ps = read_rds(atacama.rds)
```

## Check Phyloseq Object
Finally you should check that everything loaded properly with something like the following
```{r}
print(atacama.ps)
```
Which should return something like this (but with different numbers):

```
phyloseq-class experiment-level object
otu_table()   OTU Table:         [ 3388 taxa and 68 samples ]
sample_data() Sample Data:       [ 68 samples by 22 sample variables ]
tax_table()   Taxonomy Table:    [ 3388 taxa by 7 taxonomic ranks ]
```

```{r}
plot_bar(atacama.ps)

```
```{r}
plot_bar(atacama.ps, fill="Kingdom")
```

```{r}
plot_bar(atacama.ps, fill="Phylum")
```

```{r}
p = plot_bar(atacama.ps, fill="Phylum")
p + geom_bar(aes(color=Phylum, fill=Phylum), stat="identity", position="stack")

```

```{r eval=FALSE, include=FALSE}
# atacama.ps
sample_data(atacama.ps)
```

```{r}
plot_bar(atacama.ps, facet_grid=~SpeciesName, fill="Kingdom")
```

```{r}
plot_bar(atacama.ps, facet_grid=~TransectName, fill="Kingdom")
```

```{r}
plot_bar(atacama.ps, x="SpeciesName", fill="Kingdom")
```

```{r}
plot_bar(atacama.ps, x="TransectName", fill="Kingdom")
```
```{r}
plot_bar(atacama.ps, facet_grid=~TransectName, fill="Kingdom")
```

```{r}
p = plot_bar(atacama.ps, facet_grid=~TransectName, fill="Kingdom")
p + geom_bar(aes(color=Kingdom, fill=Kingdom), stat="identity", position="stack")
```



```{r}
plot_bar(atacama.ps, facet_grid=~TransectName, fill="Kingdom")
```


# Fixing some annoyances
```{r}
colnames(tax_table(atacama.ps))
tax_table(atacama.ps)[1,]

colnames(tax_table(atacama.ps)) = c("Kingdom","Phyllum", "Class","Order","Family","Genus","Species")
colnames(tax_table(atacama.ps))
```
```{r}
plot_bar(atacama.ps, fill="Kingdom")

```




```{r}
plot_bar
```

```{r}
atacama.ps.melt = psmelt(atacama.ps)
p = ggplot(atacama.ps.melt, aes_string(x = "Sample", y = "Abundance"))
p = p + geom_bar(stat = "identity", position = "stack")
p = p + theme(axis.text.x = element_text(angle = -90, hjust = 0))
p = p + theme(panel.background = element_blank())
print(p)
```

```{r}
atacama.ps.melt = psmelt(atacama.ps)
p = ggplot(atacama.ps.melt, aes_string(x = "Sample", y = "Abundance", fill="Kingdom"))
p = p + geom_bar(stat = "identity", position = "stack")
p = p + theme(axis.text.x = element_text(angle = -90, hjust = 0))
p = p + theme(panel.background = element_blank())
print(p)
```
```{r}
atacama.ps.melt = psmelt(atacama.ps)
p = ggplot(atacama.ps.melt, aes_string(x = "Sample", y = "Abundance", fill="Kingdom"))
p = p + geom_bar(stat = "identity", position = "stack")
p = p + theme(axis.text.x = element_text(angle = -90, hjust = 0))
p = p + facet_grid(~SpeciesName, scale="free_x", space = "free_x")
print(p)
```
```{r}
p = plot_bar(atacama.ps, fill="Kingdom")
p + geom_bar(aes(color=Kingdom, fill=Kingdom), stat="identity", position="stack")
```

```{r}
p = plot_bar(atacama.ps, fill="Kingdom")
p + 
  geom_bar(aes(color=Kingdom, fill=Kingdom), stat="identity", position="stack") +
  facet_grid(~SpeciesName)
```

```{r}
p = plot_bar(atacama.ps, fill="Kingdom")
p + 
  geom_bar(aes(color=Kingdom, fill=Kingdom), stat="identity", position="stack") +
  facet_grid(~SpeciesName, scale="free_x", space = "free_x")
```




