---
title: "Untitled"
output: html_document
---


```{r libraries, message=FALSE, warning=FALSE}
library(readr)
library(phyloseq)
library(tibble)
library(dplyr)
library(stringr)
library(tidyr)
library(fs)
```

```{r files_and_directories, include=FALSE}
# Directories
atacama.ps.rds = "/data/tutorial_data/atacama_1pct.rds"
outdir=path.expand("~/scratch/lefse")
# lefse.input.file=file.path(outdir,"data_for_lefse.tsv")



atacama.ps.rds %>%
  path_file %>%
  path_ext_remove %>%
  path(outdir, .) ->
  output_basename 

lefse.input.file = path_ext_set(output_basename, ".tsv")
lefse.bar = path_ext_set(output_basename, ".png")


if(dir_exists(outdir)){
  dir_delete(outdir)
}
dir.create(outdir)


Sys.setenv(LEFSE_INPUT_FILE=lefse.input.file)
Sys.setenv(OUTDIR=outdir)
Sys.setenv(BASENAME=output_basename)

# NORMALIZATION=""
Sys.setenv(NORMALIZATION="-o 1000000")
Sys.setenv(PLOT_FORMAT="png")

atacama.ps = read_rds(atacama.ps.rds)
print(atacama.ps)
```


```{r}
#' RepseqToTaxa:
#' Convert Repseq column names to Taxa column names in a spread data frame
#' The big problem here is that this needs to be done after all other 
#' manipulations to the dataframe, otherwise most functions will balk if there
#' are dataframe columns with identical names
#'
#' @param spread.df The dataframe generated from phyloseq object.
#' @param source.ps phyloseq object from which spread.df was derived.
RepseqToTaxa <- function(spread.df, source.ps) {
  tax.df = as.data.frame(tax_table(source.ps)) %>%
    rownames_to_column("repseq") %>%
    mutate(taxonomy = paste(Kingdom,Phylum,Class,Order,Family,Genus, sep="|")) %>%
    select(repseq, taxonomy)
  
  # need to preserve non-OTU column names (otherwise they get lost)
  colname_match = match(names(spread.df), tax.df$repseq)
  cols_to_keep = which(is.na(colname_match))
  colnames_to_keep = names(spread.df)[cols_to_keep]
  
  # replace repseqs with taxonomies
  names(spread.df) = tax.df$taxonomy[match(names(spread.df), tax.df$repseq)]
  # now reset the non-OTU column names
  names(spread.df)[cols_to_keep] = colnames_to_keep
  return(spread.df)
}

GenerateLefseOutput = function(ps,output_columns,outfile){
  if(length(output_columns)==1){
    format.flags="-f c -u 1 -c 2"
  }else if(length(output_columns)==2){
    format.flags="-f c -u 1 -c 2 -s 3"
  }else{
    stop("output_columns must be a vector of length 1 or 2")
  }
  base_columns = c("SampleID","OTU","Abundance")
  spread.df = psmelt(ps) %>% 
    mutate(SampleID = str_replace(Sample, pattern="\\.", replacement="_")) %>%
    mutate(taxonomy = paste(Kingdom,Phylum,Class,Order,Family,Genus, sep="|")) %>%
    select(one_of(c(base_columns,output_columns))) %>%
    spread(OTU, Abundance)
  
  RepseqToTaxa(spread.df, ps) %>%
    write.table(file=outfile, 
                sep="\t", quote = FALSE,
                row.names = FALSE)
  
  return(format.flags)
}
```

```{r}
sample_variables(atacama.ps)
```

```{r}
# format.flags = GenerateLefseOutput(atacama.ps, c("Vegetation","TransectName"), lefse.input.file)
format.flags = GenerateLefseOutput(atacama.ps, "Vegetation", lefse.input.file)
Sys.setenv(FORMAT_FLAGS=format.flags)
```


# Run With Normalization=1000000

```{bash}
format_input.py $LEFSE_INPUT_FILE  "${BASENAME}.in" $FORMAT_FLAGS $NORMALIZATION  --output_table ${BASENAME}.tab
```
# head ${BASENAME}.in ${BASENAME}.tab

# run_lefse.py performs the actual statistica analysis
#
# Apply LEfSe on the formatted data producing the results (to be further processed
# for visualization with the other modules). The option available
# can be listed using the -h option 
```{bash}
run_lefse.py "${BASENAME}.in" ${BASENAME}.res
```

# plot_res.py visualizes the output
#
# Plot the list of biomarkers with their effect size
# Severak graphical options are available for personalizing the output
# ../plot_res.py hmp_aerobiosis_small.res hmp_aerobiosis_small.png
```{bash}
plot_res.py --format ${PLOT_FORMAT} ${BASENAME}.res ${BASENAME}.${PLOT_FORMAT}
```

# plot_cladogram.py visualizes the output on a hierarchical tree
#
# Plot the representation of the biomarkers on the hierarchical tree
# specified in the input data (using | in the name of the features)
# In this case we will obtain the RDP taxonomy.
# This is an early implementation of the module. I'm working on an improved version
# that will be released independently from LEfSe
```{bash}
plot_cladogram.py ${BASENAME}.res ${BASENAME}.cladogram.${PLOT_FORMAT} --format ${PLOT_FORMAT}
```

# Create a directory for storing the raw-data representation of the discovered biomarkers
```{bash}
mkdir -p ${BASENAME}_biomarkers_raw_images
```

# plot_features.py visualizes the raw-data features
#
# The module for exporting the raw-data representation of the features.
# With the default options we will obtain the images for all the features that are
# detected as biomarkers
```{bash}
plot_features.py ${BASENAME}.in ${BASENAME}.res ${BASENAME}_biomarkers_raw_images/
```

```{r results="asis"}
cat(paste0("![lefse results](", lefse.bar, ")"), fill = FALSE)
```

