---
title: "Paths in R, RStudio, and Unix"
output:
  md_document:
    variant: markdown_github
---

# Unix
## List Files and Directories
```{bash}
ls
```

```{bash}
ls -a
```

```{bash}
ls -F
```


## Working Directory
```{bash}
pwd
```

```{bash}
cd ..
pwd
```

```{bash}
pwd
```

## Absolute vs Relative Paths
```{bash}
head paths.Rmd
```

```{bash}
head /home/guest/IBIEM_2018_2019/content/misc/paths.Rmd
```

## Environment Variables
```{bash}
echo "Hello world"
```

```{bash}
echo $HOME
```

```{bash}
echo $PWD
```

```{bash}
ls $HOME/IBIEM_2018_2019
```

## Tilde Expansion

```{bash}
echo ~
```

```{bash}
ls ~
```

```{bash}
cd ~
pwd
```

## Make a New Directory
```{bash}
mkdir bash_test_dir
ls -F
```



# R

## List Files and Directories
```{r}
list.files()
```

```{r}
list.dirs()
```

## Working Directory
```{r}
getwd()
```

## Environment Variables
```{r}
Sys.getenv()
```

```{r}
Sys.getenv("HOME")
```

```{r}
Sys.getenv("PWD")
```

## Path Expansion and Normalization
```{r}
path.expand("~////")
```

```{r}
normalizePath("~////")
```

```{r}
normalizePath("~/not_a_directory")
```

```{r}
path.expand("~/not_a_directory")
```

```{r}
path.expand("./paths.Rmd")
```

```{r}
normalizePath("./paths.Rmd")
```

## Path Maniputation

```{r}
file.path("some_dir", "some_subdir")
```

```{r}
file.path(Sys.getenv("HOME"), "IBIEM_2018_2019")
```

```{r}
this.file = "/home/guest/IBIEM_2018_2019/content/misc/paths.Rmd"
basename(this.file)
```
```{r}
dirname(this.file)
```

## Make a New Directory
```{r}
dir.create("r_test_dir")
list.dirs()
```

# RStudio
- Home
- Project Directory (R cube)
- ".."
- "..."
- Git Pane



# References
- [Navigating Files and Directories](http://swcarpentry.github.io/shell-novice/02-filedir/index.html)
- [Working With Files and Directories](http://swcarpentry.github.io/shell-novice/03-create/index.html)
